Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-07-09 semiring.tex
Sections labels et semirings

% !TEX root = main.tex
%
% Semiring basics
%
We shall consider semirings for the weight values of our language models.
%\florent{The results are established for a general class of semirings. They can be instantiated for concrete cases}
A \emph{semiring} $\< \Semiring, \oplus, \zero, \otimes, \one>$
is a structure with a domain~$\Semiring$,
equipped with two associative
binary operators~$\oplus$ and $\otimes$,
with respective neutral elements $\zero$ and $\one$, and such that:
%$\< \mathbb{S}, \oplus, \zero>$ is a commutative monoid
%$\< \mathbb{S}, \otimes, \one>$ is a monoid
\begin{itemize}
\item $\oplus$ is commutative:
 $\< \Semiring, \oplus, \zero>$ is a commutative monoid
   and $\< \Semiring, \otimes, \one>$ a monoid,
\item $\otimes$ distributes over~$\oplus$:  $\forall x, y, z \in \mathbb{S}$,
$x \otimes (y \oplus z) = (x \otimes y) \oplus (x \otimes z)$,
and $(x \oplus y) \otimes z = (x \otimes z) \oplus (y \otimes z)$,
\item $\zero$ is absorbing for~$\otimes$:
$\forall x\in \mathbb{S}$, $\zero \otimes x = x \otimes \zero = \zero$.
\end{itemize}
%Components of a semiring~$\Semiring$ may be subscripted by~$\Semiring$ when needed.
%We simply write $x \in \Semiring$ to mean $x \in \mathbb{S}$.
%
Intuitively, in the models presented in this paper,
$\oplus$ selects an optimal value from two given values,
in order to handle non-determinism,
and $\otimes$ combines two values into a single value.
%and let $\< \Semiring, \oplus, \zero, \otimes, \one>$ be a {semiring},

\medskip%\noindent
A semiring $\Semiring$ is \emph{commutative} if $\otimes$ is commutative.
It is \emph{idempotent} if for all $x \in \Semiring$, $x \oplus x = x$.
%
Every idempotent semiring~$\Semiring$ induces
a partial ordering~$\leq_\oplus$
called the \emph{natural ordering} of~$\Semiring$~\cite{Mohri02semiring}
defined,  by:
%implicitly defined by the semiring $\Semiring$
for all $x, y\in \Semiring$,
$x \leq_\oplus y \;\mbox{iff}\; x \oplus y = x$.
%(see~\cite{Mohri02semiring} for the proof that it is an ordering).
%
The natural ordering is sometimes defined in the opposite direction~\cite{DrosteKuich09semirings};
We follow here the direction  %follows \cite{Mohri02semiring}, and
that coincides with the usual ordering on the Tropical semiring \emph{min-plus}
(Figure~\ref{fig:semirings}).
%
\noindent
An idempotent semiring $\Semiring$~is called \emph{total} if
it~$\leq_\oplus$ is total
\ie when for all $x, y \in \Semiring$, either $x \oplus y = x$ or $x \oplus y = y$.
\florent{is total necessary?}

\begin{lemma}[Monotony, \cite{Mohri02semiring}] \label{lem:monotonic}
Let $\< \Semiring, \oplus, \zero, \otimes, \one>$ be an idempotent semiring.
For all $x, y, z  \in \Semiring$,
if $x \leq_\oplus y$ then
$x \oplus z \leq_\oplus y \oplus z$,
$x \otimes z \leq_\oplus y \otimes z$
and $z \otimes x \leq_\oplus z \otimes y$.
\end{lemma}
%When  holds,
%We say in this case that
To express the property of Lemma~\ref{lem:monotonic}, we call
$\Semiring$ \emph{monotonic} \wrt $\leq_\oplus$.
%A semiring $\Semiring$
%is \emph{monotonic} \wrt a partial ordering~$\leq$
%iff for all $x, y, z  \in \Semiring$,  $x \leq y$ implies
%$x \oplus z \leq y \oplus z$,
%$x \otimes z \leq y \otimes z$,
%and $z \otimes x \leq z \otimes y$.
%
Another important semiring property in the context of optimization
is {superiority}~\cite{Huang08advanceddynamic},
which corresponds to the
\emph{non-negative weights} condition in shortest-path algorithms~\cite{Dijkstra59anote}.
Intuitively, it means that combining elements with $\otimes$ always increase their weight.
Formally, it is defined as the property ($i$) below. %of the following lemma.

@@

%\medskip
\noindent
\philippe{Here the difference between $\Semiring$ as a structure and as a domain is blurred.}
We shall need below infinite sums with~$\oplus$.
A semiring~$\Semiring$ is called \emph{complete}~\cite{Droste09handbook}
%(\cite{Droste09handbook} chapter 1) %\cite{Kuich97semirings}
if it has an %infinite sum
operation $\bigoplus_{i \in I} x_i$
for every family
$(x_i)_{i \in I}$ %$\{ x_i \mid i \in I \}$
of elements of $\dom(\Semiring)$ over an index set $I \subset \mathbb{N}$, such that:
\philippe{$j\in \mathbb{N}$: $j$ is en element of $\mathbb{N}$, not the same s $j\subset \mathbb{N}$}
%the following holds: %properties
\begin{description}
\item[$i.$]
\emph{infinite sums extend finite sums:}\\
$\displaystyle\bigoplus_{i \in \emptyset} x_i = \zero$,\quad
      $\forall j\in \mathbb{N}, \displaystyle\bigoplus_{i \in \{ j \}} x_i = x_j$,
      $\forall j, k\in \mathbb{N}, j\neq k,
      \displaystyle\bigoplus_{i \in \{ j, k \}} x_i = x_j \oplus x_k$,
%
\item[$ii.$]
\emph{associativity and commutativity:}\\
for all $I \subseteq \mathbb{N}$
and all partition $(I_{j})_{j \in J}$ of $I$, %\subseteq \mathbb{N}$,
\(
\displaystyle
\bigoplus_{j \in J}\bigoplus_{i \in I_j} x_i =
\bigoplus_{i \in I} x_i
\),
%
\item[$iii.$]
\emph{distributivity of product over infinite sum:}\\
for all $I \subseteq \mathbb{N}$,
\(
\displaystyle
\bigoplus_{i \in I} (x \otimes y_i) = x \otimes \bigoplus_{i\in I} y_i\), and
\(
\displaystyle
\bigoplus_{i \in I} (x_i \otimes y) = (\bigoplus_{i \in I} x_i ) \otimes y\).
\end{description}

@@

\begin{example}\label{rules-division}
The recursive subdivision of time that leads to hierarchichal
structures of music notation can be modeled as
production rules. Since there exists several possible division,
rules can be weighted
in the tropical semiring whose domain $\mathbb{R}_+ \cup \{ +\infty\}$,
$\oplus$ is min, $\zero = +\infty$, $\otimes$ is sum, and $\one = 0$.
For instance, the following production rules define two possible divisions
of a bounded time interval into respectively a duplet and a triplet.
\[
\rho_1: q_0 \xrightarrow{0.06} \< q_{1}, q_{2}>,\
\rho_2: q_0 \xrightarrow{0.12} \< q_{1}, q_{2}, q_{2}>.
\]
Further binary divisions of time sub-intervals %represented respectively by~$q_2$ and~$q_3$
are possible with:

\[
\rho_3:\,q_2 \xrightarrow{0.1} \< q_{3}, q_{3}>, \
\rho_4:\,q_3 \xrightarrow{0.11} \< q_{4}, q_{4}>. \
\]

\endex
\end{example}
%%% semiring properties used in paper
% - commutative
% - bounded (implies idempotent and superior)
% - total
% - complete

>>>>

% !TEX root = main.tex
%
% Semiring basics
%%
%We shall consider semirings for the weight values of our language models.
%\florent{The results are established for a general class of semirings. They can be instantiated for concrete cases}
A \emph{semiring} $\< \Semiring, \oplus, \zero, \otimes, \one>$
is a structure with a domain~$\Semiring$,
equipped with two associative
binary operators~$\oplus$ and $\otimes$,
with respective neutral elements $\zero$ and $\one$, and such that:
%$\< \mathbb{S}, \oplus, \zero>$ is a commutative monoid
%$\< \mathbb{S}, \otimes, \one>$ is a monoid
\begin{itemize}
\item $\oplus$ is commutative:
 $\< \Semiring, \oplus, \zero>$ is a commutative monoid
   and $\< \Semiring, \otimes, \one>$ a monoid,
\item $\otimes$ distributes over~$\oplus$:  $\forall x, y, z \in \mathbb{S}$,
$x \otimes (y \oplus z) = (x \otimes y) \oplus (x \otimes z)$,
and $(x \oplus y) \otimes z = (x \otimes z) \oplus (y \otimes z)$,
\item $\zero$ is absorbing for~$\otimes$:
$\forall x\in \mathbb{S}$, $\zero \otimes x = x \otimes \zero = \zero$.
\end{itemize}
%Components of a semiring~$\Semiring$ may be subscripted by~$\Semiring$ when needed.
%We simply write $x \in \Semiring$ to mean $x \in \mathbb{S}$.
%
Intuitively, in the models presented in this paper,
$\oplus$ selects an optimal value from two given values,
in order to handle non-determinism,
and $\otimes$ combines two values into a single one.
%and let $\< \Semiring, \oplus, \zero, \otimes, \one>$ be a {semiring},

A semiring $\Semiring$ is \emph{commutative} if $\otimes$ is commutative.
It is \emph{idempotent} if for all $x \in \Semiring$, $x \oplus x = x$.
%
Every idempotent semiring~$\Semiring$ induces
a partial ordering~$\leq_\oplus$
called the \emph{natural ordering} of~$\Semiring$~\cite{Mohri02semiring}
defined,  by:
%implicitly defined by the semiring $\Semiring$
for all $x, y\in \Semiring$,
$x \leq_\oplus y \;\mbox{iff}\; x \oplus y = x$.
%(see~\cite{Mohri02semiring} for the proof that it is an ordering).
%
The natural ordering is sometimes defined in the opposite direction~\cite{DrosteKuich09semirings};
We follow here the direction  %follows \cite{Mohri02semiring}, and
that coincides with the usual ordering on the Tropical semiring \emph{min-plus}
(Figure~\ref{fig:semirings}).
%
\noindent
An idempotent semiring $\Semiring$~is called \emph{total} if
it~$\leq_\oplus$ is total
\ie when for all $x, y \in \Semiring$, either $x \oplus y = x$ or $x \oplus y = y$.
\florent{is total necessary?}

\begin{lemma}[Monotony, \cite{Mohri02semiring}] \label{lem:monotonic}
Let $\< \Semiring, \oplus, \zero, \otimes, \one>$ be an idempotent semiring.
For all $x, y, z  \in \Semiring$,
if $x \leq_\oplus y$ then
$x \oplus z \leq_\oplus y \oplus z$,
$x \otimes z \leq_\oplus y \otimes z$
and $z \otimes x \leq_\oplus z \otimes y$.
\end{lemma}
%When  holds,

We then say the
$\Semiring$ is \emph{monotonic} \wrt $\leq_\oplus$.
%A semiring $\Semiring$
%is \emph{monotonic} \wrt a partial ordering~$\leq$
%iff for all $x, y, z  \in \Semiring$,  $x \leq y$ implies
%$x \oplus z \leq y \oplus z$,
%$x \otimes z \leq y \otimes z$,
%and $z \otimes x \leq z \otimes y$.
%
Another important semiring property in the context of optimization
is {superiority}~\cite{Huang08advanceddynamic},
which corresponds to the
\emph{non-negative weights} condition in shortest-path algorithms~\cite{Dijkstra59anote}.
Intuitively, it means that combining elements with $\otimes$ always increase their weight.
Formally, it is defined as the property ($i$) below. %of the following lemma.

@@

%\medskip
\noindent
We  need infinite sums with~$\oplus$.
A semiring~$\Semiring$ is called \emph{complete}~\cite{Droste09handbook}
%(\cite{Droste09handbook} chapter 1) %\cite{Kuich97semirings}
if it has an %infinite sum
operation $\bigoplus_{i \in I} x_i$
for every family
$(x_i)_{i \in I}$ %$\{ x_i \mid i \in I \}$
of elements of $\dom(\Semiring)$ over an index set $I \subset \mathbb{N}$, such that:
\begin{description}
\item[$i.$]
\emph{infinite sums extend finite sums:}\\
$\displaystyle\bigoplus_{i \in \emptyset} x_i = \zero$,\quad
      $\forall j\in \mathbb{N}, \displaystyle\bigoplus_{i \in \{ j \}} x_i = x_j$,
      $\forall j, k\in \mathbb{N}, j\neq k,
      \displaystyle\bigoplus_{i \in \{ j, k \}} x_i = x_j \oplus x_k$,
%
\item[$ii.$]
\emph{associativity and commutativity:}\\
for all $I \subseteq \mathbb{N}$
and all partition $(I_{j})_{j \in J}$ of $I$, %\subseteq \mathbb{N}$,
\(
\displaystyle
\bigoplus_{j \in J}\bigoplus_{i \in I_j} x_i =
\bigoplus_{i \in I} x_i
\),
%
\item[$iii.$]
\emph{distributivity of product over infinite sum:}\\
for all $I \subseteq \mathbb{N}$,
\(
\displaystyle
\bigoplus_{i \in I} (x \otimes y_i) = x \otimes \bigoplus_{i\in I} y_i\), and
\(
\displaystyle
\bigoplus_{i \in I} (x_i \otimes y) = (\bigoplus_{i \in I} x_i ) \otimes y\).
\end{description}

@@

\begin{example}\label{rules-division}
The recursive subdivision of time that leads to hierarchichal
structures of music notation can be modeled as
production rules. Since there exists several possible division,
rules can be weighted in the tropical semiring.
For instance, the following production rules define two possible divisions
of a time interval into respectively a duplet and a triplet.
\[
\rho_1: q_0 \xrightarrow{0.06} \< q_{1}, q_{2}>,\
\rho_2: q_0 \xrightarrow{0.12} \< q_{1}, q_{2}, q_{2}>.
\]
Further binary divisions of time sub-intervals %represented respectively by~$q_2$ and~$q_3$
are possible with:
\[
\rho_3:\,q_2 \xrightarrow{0.1} \< q_{3}, q_{3}>, \
\rho_4:\,q_3 \xrightarrow{0.11} \< q_{4}, q_{4}>. \
\]
\end{example}
%%% semiring properties used in paper
% - commutative
% - bounded (implies idempotent and superior)
% - total
% - complete
